---
title: "Data Visualization tutorial"
author: "by Dominic White"
tutorial:
  version: 1.3
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

<style>
p.comment {
background-color: #EDBB99;
padding: 10px;
border: 1px solid black;
font-style: italic;
}
</style>

<style>
p.comment1 {
background-color: #A2D9CE;
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment2 {
background-color: #EDBB99; 
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment3 {
background-color: #FDFEFE; 
padding: 10px;
border: 1px solid black;
}
</style>


```{r setup, include=FALSE, context="setup"}
library(tidyverse)
library(learnr)
library(gradethis)
library(checkr)
library(sortable)
library(diffobj)
library(flair)
library(listviewer)

knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)
tutorial_options(exercise.startover = TRUE)
tutorial_options(exercise.completion = FALSE)

new_recorder <- function(tutorial_id, tutorial_version, user_id, event, data) {
    #cat(tutorial_id, " (", tutorial_version, "): ", user_id, ", ", event, ", ", "\n", sep = "")
  cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
  cat("event: ", event, "\n", sep = "")
}

options(tutorial.event_recorder = new_recorder)
```

## Welcome

This tutorial will introduce you to creating graphs in R.

## Visualization principles

This video will introduce you to the importance of visualization in data science, and how it can be used effectively (or ineffectively).

<iframe width="560" height="315" src="https://www.youtube.com/embed/tHEVRahx0WE?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/ETyT0aRqt6tHspWXcGyumEgBG6K79TUuaN0mWTLsaiOySA?e=KkfEsO)


## Data Overview

Before we begin to make our own graphs, we need to learn some terminology for describing the underlying data. 

<iframe width="560" height="315" src="https://www.youtube.com/embed/zM5y3dZ-o6o?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/ETqhI64qDaFGja02KJaA_usBe0iEVyOhjM5kH7ENZswmsg?e=mg4B0G)


## Creating Basic Graphs

This video will teach you how to create simple graphs in R.

<iframe width="560" height="315" src="https://www.youtube.com/embed/N3dYCR75RqA?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/Eflj-ZbwysRIjeWbUg1wSoQBAf26ac5hF_RPRRjpylmI8A?e=9jM9YC)

### The `ggplot2` library

Although the `qplot()` function has already been defined for us, we still need to tell R where to find that definition before we can use the function.

Functions like `qplot()` are usually collected into *packages* which we can then install and load in R when we need them. Datasets can also be included in packages.

* The `qplot()` function is part of the `ggplot2` package which contains functions for creating graphs.
* The `starwars` dataset is part of the `dplyr` package which mostly contains functions for manipulating datasets (we will learn about this soon in the data wrangling module).

Before we can use `qplot()` or `starwars`, we need to import their packages. We import a package using the `library()` function.

### Load the `tidyverse` package

The `tidyverse` package actually a *meta-package*: a collection of multiple other packages that are commonly used together. `ggplot2` and `dplyr` are both part of the tidyverse. We can thus run `library(tidyverse)` to load both of these packages at once.

Load the `tidyverse` meta-package using the `library()` function with the correct argument:

```{r xxdpwjzwuhmszksa, exercise = TRUE}
# replace these underscores with the code
____
```


```{r xxdpwjzwuhmszksa-hint-1}
"Type in: library(tidyverse)"
```


```{r xxdpwjzwuhmszksa-solution}
# solution code
library(tidyverse)
```

```{r xxdpwjzwuhmszksa-check}
# check code
gradethis::grade_code()
```

### The `starwars` dataset

Now that we have loaded the `tidyverse` meta-package (and thus the `dplyr` package that it contains, we have access to the dataset of Star Wars character statistics stored in the `starwars` variable.

Run the following code chunk to inspect the dataset, and then answer the following quiz questions:

```{r osikxouifqvtzztk, exercise = TRUE}
starwars
```

`starwars` is an example of a **dataframe**: a table of data with rows and columns. Each row contains an *observation*, and each column holds a *variable*.

```{r starwars_col_quiz, echo=FALSE}
question("How many variables are in this dataset?",
  answer("14", correct = TRUE),
  answer("4"),
  answer("87"),
  answer("10"),
  answer("77"),
  allow_retry = TRUE
)
```

```{r starwars_rows_quiz, echo=FALSE}
question("How many observations are in this dataset?",
  answer("14"),
  answer("4"),
  answer("87", correct = TRUE),
  answer("10"),
  answer("77"),
  allow_retry = TRUE
)
```

```{r starwars_obs_quiz, echo=FALSE}
question("What does each observation in this dataset represent?",
  answer("A fact about a Star Wars character, e.g. their eye color or height"),
  answer("A series of data about the same characteristic, e.g. eye color"),
  answer("A single Star Wars character", correct = TRUE),
  answer('"These are not the answers you are looking for."'),
  allow_retry = TRUE
)
```


## Interpreting Scatter Plots

This video will teach you about the patterns that can be described in scatter plots.

<iframe width="560" height="315" src="https://www.youtube.com/embed/1qy42UG9KxI?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/EX4UdCpVmZ9Iovu6VUWUeUcB6MPiqerCa6cHHaDNoDPBcQ?e=bcNKiW)

### Interpret these scatter plots

Pick the correct description for each of these scatter plots:

```{r, echo = FALSE}
x <- runif(100,0,100)
y <- x * -2 + rnorm(100, 0, 5)
ggplot2::qplot(x,y)
```

```{r scatter_plot_interp_1, echo=FALSE}
question("The relationship in the scatter plot shown above is best described as ...:",
  answer("Non-linear"),
  answer("Linear and positive"),
  answer("Linear and negative", correct = TRUE),
  answer("Showing no relationship between x and y"),
  allow_retry = TRUE
)
```

```{r, echo = FALSE}
x <- runif(100,0,100)
y <- x^2 + x*rnorm(100, 0, 10)
ggplot2::qplot(x,y)
```

```{r scatter_plot_interp_2, echo=FALSE}
question("The relationship in the scatter plot shown above is best described as ...:",
  answer("Non-linear", correct = TRUE),
  answer("Linear and positive"),
  answer("Linear and negative"),
  answer("Showing no relationship between x and y"),
  allow_retry = TRUE
)
```

## Graphs of One Variable

Scatter plots show the co-variation between two variables. But often we want to examination in just a single variable.

This video will show you how to create histograms to visualize continuous variables and bar graphs to visualize categorical variables.

<iframe width="560" height="315" src="https://www.youtube.com/embed/exEQJozQNt0?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/Ec31has3P-RGnhRzZG1sCOUBN1JuMUTicFQts5kajlHvyA?e=CIMnUj)


<p class="comment3">
Fill in the blank to create a histogram of the `mass` variable from the `starwars` dataset. (You can ignore any warning about the number of bins.)
</p>

```{r gsfewpofvjsvmdfg, warning=FALSE, exercise = TRUE}
# replace the underscores with the variable
qplot(x = ____, data = starwars)
```


```{r gsfewpofvjsvmdfg-hint-1}
# hint text
"Replace the underscores with `mass`"
```


```{r gsfewpofvjsvmdfg-solution}
# solution code
qplot(x = mass, data = starwars)
```

```{r gsfewpofvjsvmdfg-check}
# check code
gradethis::grade_code()
```



## Interpreting Histograms

This video will teach you how to describe the shape of histograms.

<iframe width="560" height="315" src="https://www.youtube.com/embed/TmPShy8gK24?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://gmuedu-my.sharepoint.com/:b:/g/personal/dwhite34_gmu_edu/Ea_4rRVzhrhAnQSzQYvRrnoBrqJI-FM9FPb7Qr2RFyiFow?e=sixkdy)

### Outliers in histograms

Here is the histogram of the `mass` of Star Wars characters that you created earlier.
  
```{r, echo=FALSE, warning=FALSE}
qplot(x = mass, data = starwars)
```

```{r histogram_interp_1, echo=FALSE}
question("Is there an outlier in this histogram?",
  answer("Yes", correct = TRUE),
  answer("No"),
  allow_retry = TRUE
)
```

### A galaxy without Jabba

Here is the histogram of the `starwars` dataset characters *without* Jabba the Hutt. The number of bins has also been reduced to decrease the jaggedness of the distribution.

```{r, warning=FALSE, echo=FALSE}
qplot(x = mass, data = filter(starwars, name!="Jabba Desilijic Tiure"), bins = 7)
```

Answer the following questions about this histogram:

```{r histogram_interp_2, echo=FALSE}
question("How many bins can you see in this histogram?",
  answer("1"),
  answer("3"),
  answer("5"),
  answer("7", correct = TRUE),
  answer("8"),
  answer("9"),
  answer("10"),
  answer("15"),
  answer("20"),
  answer("30"),
  allow_retry = TRUE
)
```

```{r histogram_interp_3, echo=FALSE}
question("How would you describe the modality of this histogram?",
  answer("Unimodal", correct = TRUE),
  answer("Bimodal"),
  answer("Multi-modal"),
  answer("Uniform"),
  allow_retry = TRUE
)
```

```{r histogram_interp_4, echo=FALSE}
question("How would you describe the skewness of this histogram?",
  answer("Symmetric"),
  answer("Right skewed", correct = TRUE),
  answer("Left skewed"),
  allow_retry = TRUE
)
```
