---
title: "Advanced Statistical Inference"
author: "by Dominic White"
tutorial:
  version: 1.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

<style>
p.comment {
background-color: #EDBB99;
padding: 10px;
border: 1px solid black;
font-style: italic;
}
</style>

<style>
p.comment1 {
background-color: #A2D9CE;
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment2 {
background-color: #EDBB99; 
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment3 {
background-color: #FDFEFE; 
padding: 10px;
border: 1px solid black;
}
</style>


```{r setup, include=FALSE}
# devtools::install_github("rstudio/packrat")
library(tidyverse)
library(learnr)
library(gradethis)
library(checkr)
library(sortable)

knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)
tutorial_options(exercise.startover = FALSE)
tutorial_options(exercise.completion = FALSE)

new_recorder <- function(tutorial_id, tutorial_version, user_id, event, data) {
    #cat(tutorial_id, " (", tutorial_version, "): ", user_id, ", ", event, ", ", "\n", sep = "")
  cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
  cat("event: ", event, "\n", sep = "")
}

options(tutorial.event_recorder = new_recorder)
```

## Statistical Inference Review

### Video overview

This video reviews the inference material covered in the first module.

<iframe width="560" height="315" src="https://www.youtube.com/embed/pVAxez7pA1M?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/1uQMSTvCp1sWsH2OY1nLfUjpXaHKgRK6s/view?usp=sharing)

### Test your understanding

<!--
Other questions to add:
* What is observed statistic
* What is the null distribution
* interpreting the p-values meaning -> the probability of coming from the null
* alpha
-->

```{r response_vs_explanatory-quiz}
question(
  "You give sick patients either a drug or a placebo, and you measure whether they die or survive. This is stored in 2 variables, `treatment` and `outcome`, which are both categorical. What are your explanatory and response variables?",
  answer("response = `treatment`, explanatory = `outcome`"),
  answer("response = `outcome`, explanatory = `treatment`", correct = TRUE),
  answer('response = `"survived"`, explanatory = `"drug"`'),
  answer('response = `"survived"`, explanatory = `"placebo"`'),
  allow_retry = TRUE
)
```

```{r simulations_p_value-quiz}
question(
  "In a particular hypothesis test, we generate 10,000 permutations of our original data. 200 of them have a test statistic more extreme than our observed statistic. What is the p-value?",
  answer("0"),
  answer("0.01"),
  answer("0.02", correct = TRUE, message="The p-value is the probability that a value as or more extreme than the observed statistic came from the null distribution, which equivalent to the fraction of simulations more extreme than the observed statistic. This is 200/10,000 = 0.02"),
  answer("0.05"),
  answer("0.1"),
  answer("0.2"),
  answer("0.5"),
  answer("1"),
  answer("2"),
  answer("5"),
  answer("50"),
  allow_retry = TRUE
)
```





### Supplementary reading

These two chapters provide a good review of the original statistical inference material that we covered in the first inference module. You should review them before starting this week's assignment:

* Statistical Inference via Data Science by Chester Ismay and Albert Kim (available online at: https://moderndive.com/)
  * Chapter 7: Sampling - https://moderndive.com/7-sampling.html
  * Chapter 9: Hypothesis testing - https://moderndive.com/9-hypothesis-testing.html
  
On a lighter note, you can get an animated explanation of a hypothesis test with alpacas here: https://www.jwilber.me/permutationtest/

### Code review

In the first inference module, we used the `infer` package to simulate the null distribution.

We ran 10,000 random *permutations* to shuffle up the values of the response and explanatory variables. This random shuffling ensures that we simulate a distribution where there is no relationship between the two variables.

```r
<DATAFRAME> %>%
  specify(<RESPONSE> ~ <EXPLANATORY>) %>%
  hypothesize(null = "independence") %>%
  generate(10000, type = "permute") %>%
  calculate(stat = "...", order = c(..., ...))
```

* The `specify` function defines the response and explanatory variables.
* The `hypothesize` function declares what the null hypothesis is between those variables. `"independence"` means that there is no relationship between them.
* The `generate` function is where the simulations are actually run: we specify how many, and how to shuffle the data (in this case, using permutations).
* The `calculate` functions takes the simulations and calculates a *test statistic*: i.e. a single value for each one (e.g. the difference in proportions, or the difference in means).

<!--
### Test how much you remember!
-->

## One-sided vs. two-sided tests

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/krwRh7G6FPI?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/1EqzByiMEvi9V9MfPQ9XsLT-wcSkU69Fq/view?usp=sharing)


### 1 or 2?

<!--
Questions to add:
* Writing alternative hypotheses for 1 and 2 sides
* Which parameter in code specifies order of diff calculation -> order
* Examples of whether to use 1 or 2 sided.
* Calculating p-value
  * 1 sided left and right
  * 2 sided
-->

```{r alt_hypothesis_specification-quiz}
question(
  "You want to conduct a *two-sided* hypothesis test on data from an experiment comparing whether a patient survive a disease after receiving a drug or a placebo. What is your alternative hypothesis?",
  answer("The difference in proportions of patients who survive is significantly different between drug and placebo groups.", correct = TRUE, message = "In a two-sided test, the alternative hypothesis is that there is a difference, regardless of direction (i.e. whether the difference is positive or negative is unimportant)."),
  answer("The difference in means of patients who survive is significantly different between drug and placebo groups."),
  answer("The difference in means of patients who survive is significantly greater in the drug group than the placebo group."),
  answer("The difference in proportions of patients who survive is significantly greater in the drug group than the placebo group."),
  allow_retry = TRUE
)
```


## Effect sizes

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/o7oJPQMUS6A?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/1Ysg-oR6FLpuJ2OQ9ArzPNLvakImMgpBP/view?usp=sharing)

<!--
### Question time

-->

## Confidence intervals

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/iDxp6DHMHcc?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/11eyvzCjS_M-owi_6Ry3WH0_5sqk5Y009/view?usp=sharing)


### Questions

<!--
Extra questions?
-->

```{r permutations_vs_boostrap_1_quiz}
question(
  "You want to conduct a hypothesis test using the `infer` package in R. How should you shuffle the data to generate a null distribution?",
  answer("Bootstraps"),
  answer("Permutations", correct = TRUE, message = "A permutation shuffles up the values of the explanatory and response variables, and so simulates a world in which there is no relationship between the two (i.e. the null hypothesis)."),
  answer("It doesn't matter"),
  allow_retry = TRUE
)
```

```{r permutations_vs_boostrap_2_quiz}
question(
  "You want to calculate the confidence interval around a statistic (e.g. the mean) using the `infer` package in R. How can you approximate resampling the original population?",
  answer("Bootstraps", correct = TRUE, message = "By taking random samples of rows from the original sample (*bootstrap resampling*), we can approximate resampling the original population."),
  answer("It doesn't matter"),
  answer("Permutations"),
  allow_retry = TRUE
)
```


## Further reading

If you wish to do further reading on *bootstrapping* and *confidence intervals*, here are some online book chapters on the topic. (They cover the same material, so pick the one that makes the most sense to you.)

* Statistical Inference via Data Science by Chester Ismay and Albert Kim
  * Chapter 8: Bootstrapping and Confidence Intervals - https://moderndive.com/8-confidence-intervals.html
  
* Statistical Modeling by Daniel Kaplan
  * Chapter 5: Confidence Intervals - https://dtkaplan.github.io/SM2-bookdown/confidence-intervals.html

