---
title: "Introduction to Statistical Inference"
author: "by Dominic White"
tutorial:
  version: 1.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

<style>
p.comment {
background-color: #EDBB99;
padding: 10px;
border: 1px solid black;
font-style: italic;
}
</style>

<style>
p.comment1 {
background-color: #A2D9CE;
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment2 {
background-color: #EDBB99; 
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment3 {
background-color: #FDFEFE; 
padding: 10px;
border: 1px solid black;
}
</style>


```{r setup, include=FALSE}
# devtools::install_github("rstudio/packrat")
library(tidyverse)
library(learnr)
library(gradethis)
library(checkr)
library(sortable)

knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)
tutorial_options(exercise.startover = FALSE)
tutorial_options(exercise.completion = FALSE)

new_recorder <- function(tutorial_id, tutorial_version, user_id, event, data) {
    #cat(tutorial_id, " (", tutorial_version, "): ", user_id, ", ", event, ", ", "\n", sep = "")
  cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
  cat("event: ", event, "\n", sep = "")
}

options(tutorial.event_recorder = new_recorder)
```

<!--
TODO:
 - add in more, better quizzes and exercises throughout!
-->


## What is inference?

> Statistical inference is the process of using data analysis to deduce properties of an underlying distribution of probability. Inferential statistical analysis infers properties of a population, for example by testing hypotheses and deriving estimates.
>
> *Source: [Wikipedia](https://en.wikipedia.org/wiki/Statistical_inference)*

What does the above definition mean in simple terms?

The world around us is an uncertain, random place. Take the heights of the students in this class: you all have a wide range of heights - some taller, some shorter. 

We could plot a histogram of your heights - we would call this graph a *distribution* of your heights, and we could *infer* various things about it, e.g. the mean.

This process of going from raw measurements (e.g. of heights) to properties of the distribution that those measurements came from is called *statistical inference*.


## Distributions

### Data sampling

<iframe width="560" height="315" src="https://www.youtube.com/embed/VYChbNddUkM?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/1qInq4DXzYsHjmnw2lVu_ffVrAUuzN5ko/view?usp=sharing)

<!--
TODO:
 - Add a separate section with some additional explanation of correlation vs. causation
-->

```{r correlation_vs_causation_quiz}
question(
  "You create a linear model with a high $R^{2}$ - does this mean that:",
  answer("A change in the explanatory variable, *x*, **causes** a change in the response variable, *y*"),
  answer(
    "A change in the explanatory variable, *x*, is **correlated** with a change in the response variable, *y*", 
    correct = TRUE,
    message = "In both a linear model, and the hypothesis tests you will learn about here, we can **never** say that the explanatory variable *causes* the change in the response, only that the changes are correlated."
    ),
  allow_retry = TRUE
)
```


### Quantifying data distributions

<iframe width="560" height="315" src="https://www.youtube.com/embed/BL5Bt-nI22k?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/data_distributions_01_quantifying_data_distributions.pdf)


### Quantifying data distributions

<iframe width="560" height="315" src="https://www.youtube.com/embed/tvUw91TrWO4?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/data_distributions_02_quantifying_data_distributions_in_r.pdf)

### Create a boxplot

Fill in the ellipses (`...`) to create a boxplot of the `Sepal.Length` variable from the `iris` dataset.

```{r tfowvuehwkrgqrbv, exercise = TRUE}
# replace the three ellipses
iris %>%
  ggplot() +
  geom_...(
    mapping = aes(
      x = ...,
      y = ...
    )
  )
```

```{r tfowvuehwkrgqrbv-solution}
# solution code
iris %>%
  ggplot() +
  geom_boxplot(
    mapping = aes(
      x = "",
      y = Sepal.Length
    )
  )
```

```{r tfowvuehwkrgqrbv-check}
# check code
gradethis::grade_code()
```


### Probability mass functions

<iframe width="560" height="315" src="https://www.youtube.com/embed/xbLim5imqe4?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/data_distributions_03_probability_mass_functions.pdf)


## Hypothesis testing

### Gender discrimination case study

<iframe width="560" height="315" src="https://www.youtube.com/embed/EL9bbzblRq8?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/inference_and_simulation_01_gender_discrimination_case_study.pdf)


### A hypothesis test as a court trial

<iframe width="560" height="315" src="https://www.youtube.com/embed/LzsLhuxYiMs?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/inference_and_simulation_02_a_hypothesis_test_as_a_court_trial.pdf)


### A manual simulation of the gender discrimination experiment

<iframe width="560" height="315" src="https://www.youtube.com/embed/hjVjAzuMRDM?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/inference_and_simulation_03_a_manual_simulation_of_the_gender_discrimination_experiment.pdf)


### Simulating the gender discrimination experiment in R

<iframe width="560" height="315" src="https://www.youtube.com/embed/8wQ-uU611vI?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](http://sp19-001.cds101.com/doc/inference_and_simulation_04_simulating_the_gender_discrimination_experiment_in_r.pdf)


### One-sided hypothesis tests using infer

<iframe width="560" height="315" src="https://www.youtube.com/embed/YmeQ9ThQw9M?rel=0" frameborder="0" allowfullscreen></iframe>

Slides: [PDF](https://drive.google.com/file/d/1SUewphETMnIvzhbXoKK5mXQXhRxpLxhv/view?usp=sharing)









