---
title: "Modeling for Prediction"
author: "by Dominic White"
tutorial:
  version: 1.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

<style>
p.comment {
background-color: #EDBB99;
padding: 10px;
border: 1px solid black;
font-style: italic;
}
</style>

<style>
p.comment1 {
background-color: #A2D9CE;
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment2 {
background-color: #EDBB99; 
padding: 10px;
border: 1px solid black;
}
</style>

<style>
p.comment3 {
background-color: #FDFEFE; 
padding: 10px;
border: 1px solid black;
}
</style>


```{r setup, include=FALSE}
# devtools::install_github("rstudio/packrat")
library(tidyverse)
library(learnr)
library(gradethis)
library(checkr)
library(sortable)

knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)
tutorial_options(exercise.startover = FALSE)
tutorial_options(exercise.completion = FALSE)

new_recorder <- function(tutorial_id, tutorial_version, user_id, event, data) {
    #cat(tutorial_id, " (", tutorial_version, "): ", user_id, ", ", event, ", ", "\n", sep = "")
  cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
  cat("event: ", event, "\n", sep = "")
}

options(tutorial.event_recorder = new_recorder)
```

## Predictive modeling

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/HwZwWkxqafs?rel=0" frameborder="0" allowfullscreen></iframe>

Link to slides: [PDF](https://drive.google.com/file/d/1ui9_U0d0lgNyl1psulAigETQPV5UDdv8/view?usp=sharing)

### Test yourself

```{r terminology-1_quiz}
question(
  "In a predictive model, what is a feature?",
  answer("An explanatory variable, $x$", correct = TRUE),
  answer("The response variable, $y$"),
  answer("An observation (i.e. a row in the dataset)"),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```



## Data splits

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/SOvBYm9wyuA?rel=0" frameborder="0" allowfullscreen></iframe>

Link to slides: [PDF](https://drive.google.com/file/d/19GhjI_D0CSFiBjG7yYcxQse81T9EUiU6/view?usp=sharing)


## Logistic regression

### Video overview

<iframe width="560" height="315" src="https://www.youtube.com/embed/quKGW2Lltt0?rel=0" frameborder="0" allowfullscreen></iframe>

Link to slides: [PDF](https://drive.google.com/file/d/12V_i-IHLMI32UQMJoLwVRrCmgt_6_h0w/view?usp=sharing)

### Test yourself

```{r glm_quiz}
question(
  "What type of link function should you use to map a linear model to a binary response (i.e. logistic regression)? I.e. what is the `family` argument of the `glm()` function?",
  answer("binary"),
  answer("binomial", correct = TRUE),
  answer("gaussian"),
  answer("poisson")
)
```


